\emph{icepack}: a novel glacier flow modeling package

### 2019-08-15 icepack.tex
Add MISMIP+ demo

\begin{abstract}
In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for the describing of weak forms of partial differential equations.
The diagnostic models in icepack are formulated through action principles that are specified in UFL.
The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
Many post-processing and analysis tasks on simulation results also amount to the evaluation of some functional.
By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
A third advantage of variational principles is that the action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of extra computational expense over 2D, depth-averaged models.
Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
\end{abstract}

@@

Numerical modeling has become an essential part of the workflow of glaciologists across all disciplines.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like.
\end{enumerate}
Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several glacier flow models already exist and are effective in the hands of experts.
These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in numerical modeling.
Our goal is to make a tool that will be both accessible to non-experts and useful for experts.

\textcolor{red}{We focus in this paper only on novel aspects of icepack that differ substantially from existing packages.}

The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for specifying weak forms of PDE.
Icepack improves upon the groundwork laid in VarGlaS through its use of tensor product finite elements on extruded meshes, both of which are only available in firedrake \citep{bercea2016structure, mcrae2016automated}.
These two features enable the definition of a much simpler 3D flow model that we will describe in the next section.

@@

\section{Diagnostic models}

There are two diagnostic models implemented in icepack.
The \emph{shallow stream approximation} (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
The \emph{first-order} or \emph{Blatter-Pattyn} (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

Both the SSA and BP models in icepack are described through \emph{action principles} \citep{dukowicz2010consistent}.
Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a certain nonlinear functional, called the action.
The action consists of four terms:
\begin{align}
    & \text{action} = \iint\text{stress} \times \text{strain rate}\ud z\ud x - \int\text{basal friction} \times \text{sliding velocity}\ud x \nonumber \\
    & \quad - \iint\text{surface slope}\times\text{velocity}\ud z\ud x - \iint\text{ocean pressure}\times\text{velocity}\ud z\ud \gamma
    \label{action-functional}
\end{align}
where $\ud z\ud x$ denotes integration over the entire glacier, $\ud x$ denotes integration over the glacier footprint, and $\ud z\ud\gamma$ over the side wall boundary.
The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
Moreover, the energy lost to viscous and frictional heating can be calculated from the action and its Legendre transform.
The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium \citep{de2013non}.

@@

\subsection{Implementation with firedrake}

At the highest level, each physics model in icepack is represented by a Python class.
So far, the models available are \texttt{IceShelf}, \texttt{IceStream}, and \texttt{HybridModel}.
Each class has three main responsibilities.
First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.

\textcolor{red}{Some sample code.}

The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
These primitives largely consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.

\textcolor{red}{Brief discussion of SSA model.}

\subsection{Terrain-following coordinates and extruded meshes}

The 3D flow model, as represented in the \texttt{HybridModel} class, uses several features that are only available in Firedrake to better exploit the underlying mathematics.
Rather than represent the 3D physical domain itself, we instead use \emph{terrain-following coordinates} and \emph{extruded meshes}.
The terrain-following vertical coordinate $\zeta$ is defined as
\begin{equation}
    \zeta = \frac{z - b}{h}
\end{equation}
where $b$ is the ice base.
We can then think of the computational domain as the Cartesian product of a 2D footprint domain $\Omega$ and the unit interval $[0, 1]$.

Both the bed elevation and thickness depend on $x$ and $y$.
As a result, the formula for the horizontal gradient of a field in terrain-following coordinates includes an additional geometric correction factor.
Letting $\nabla_z$ and $\nabla_\zeta$ denote the horizontal gradient with respectively $z$ and $\zeta$ held constant, the chain rule gives us that
\begin{equation}
    \nabla_zq = \nabla_\zeta q + \frac{\partial q}{\partial\zeta}\nabla\zeta,
\end{equation}
where we can calculate the spatial gradient of $\zeta$ as
\begin{equation}
    \nabla\zeta = -h^{-1}\left\{(1 - \zeta)\nabla b + \zeta\nabla s\right\}.
\end{equation}
Likewise, the strain rate of a vector field can be expressed as
\begin{equation}
    \dot\varepsilon_z(u) = \dot\varepsilon_\zeta(u) + \frac{1}{2}\left(u \otimes\nabla\zeta + \nabla\zeta\otimes u\right)
\end{equation}
where $\otimes$ is the tensor product of two vectors.
The geometric correction factor can easily be represented in UFL.
By defining a wrapper around the UFL \texttt{grad} function, the code to define the action functional in terrain-following coordinates is no more complex than in Cartesian coordinates.

Terrain-following coordinates are especially advantageous on extruded meshes.
An extruded mesh takes a 2D footprint mesh and lifts it into 3D by taking the Cartesian product with the unit interval.
The cells of an extruded mesh are triangular prisms instead of the more common tetrahedra used for general 3D meshes.
In Firedrake, an extruded mesh is created by calling the function \texttt{ExtrudedMesh} on the 2D footprint.
Not every 3D domain can be described by extruding a 2D domain, but the geometry of most glacier flow problems can.

@@

We also argue that answers expressed in terrain-following coordinates are more intuitive in some respects than in Cartesian coordinates.
At the bed of a grounded glacier, the vertical velocity in Cartesian coordinates is
\begin{equation}
    w = -\dot a_b + u\cdot\nabla b.
\end{equation}
Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
In other words it is not immediately clear whether the vertical velocity is more a result of basal mass balance or of geometry without more information.
By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.

\subsection{Vertical spectral elements}

Having chosen to use an extruded mesh, we must then decide how many vertical layers to use.
In Elmer/Ice, for example, the number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.
The horizontal velocity for many realistic flows is very smooth as a function of depth.
For example, under the plug flow approximation, the horizontal velocity is constant with depth.
Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
This extra information about our solution suggests a modal rather than a nodal discretization strategy.

Rather than extrude the footprint mesh into many vertical layers, we might instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
To support this use case, Firedrake includes the ability to define \emph{tensor product} elements on extruded meshes.
Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
\begin{equation}
    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
\end{equation}
For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
Tensor product elements are defined in Firedrake by passing the extra keyword arguments \texttt{vfamily}, \texttt{vdegree} to the constructor for a function space.
Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
To select the Legendre polynomial basis, the user passes the keyword argument \texttt{vfamily=`Gauss-Legendre'} or \texttt{`GL'} for short to the constructor for the function space.

The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
This approach can be thought of merely as a way to discretize a PDE that has special structure.
Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
The system is then discretized in the horizontal and solved numerically.
In any case, the code is the same regardless of how one wishes to think about the underlying mathematics.

@@

This approach for implementing a 3D flow model works completely seamlessly but for one important detail.
The backpressure from ocean water at the calving front of a marine-terminating glacier is not a smooth function of depth.
The pressure is 0 above the water line and linearly increasing below it:
\begin{equation}
    \text{backpressure power} = \int_\Gamma\int_0^1 \rho_Wgh(\zeta_{\text{sl}} - \zeta)_+\ud\zeta\ud\gamma,
    \label{backpressure}
\end{equation}
where $\zeta_{\text{sl}}$ denotes the relative depth to the water line and the subscript $+$ denotes the positive part of a real number.
Were we to use the standard asssembly procedure in Firedrake to evaluate this integral, we would get an inaccurate result due to an insufficient number of integration points.
The resulting velocity solutions are then wildly inaccurate due to the mis-specification of the Neumann boundary condition.
A blunt solution to this problem would be to pass an extra argument to the Firedrake form compiler that specifies a much greater integration accuracy in the vertical for this term.
This fix reduces the errors in the velocities, but it does not eliminate them completely and it incurs a large computational cost.

We instead implemented a routine that symbolically calculates the Legendre polynomial expansion of the function $(\zeta_{\text{sl}} - \zeta)_+$ with respect to the parameter $\zeta_{\text{sl}}$ using the package SymPy \citep{sympy}.
The symbolic variables for $\zeta$ and $\zeta_{\text{sl}}$ used in the SymPy representation of the polynomial expansion are then substituted for equivalent symbolic variables in Firedrake/UFL using the SymPy object's \texttt{subs} method.
The Legendre polynomial approximation to this function only converges linearly as the number of coefficients is increased, since the the function is continuous but not smooth, and the approximation exhibits noticeable ringing artifacts \textcolor{red}{figure}.
While the approximation itself is not very accurate, the calculated value of the integral in equation \eqref{backpressure} is exact because of the orthogonality property.
Stated another way, the residuals in the approximation are large, but they integrate to 0 when multiplied by any Legendre polynomial up to the number of vertical modes.

\textcolor{red}{Speed of hybrid model.}
One of the main advantages of this hybrid model is the ease with which a more accurate solution can be bootstrapped from a less accurate solution, since the two fields are defined on exactly the same geometry.
For example, if we only have a very crude initial guess for the solution, we can obtain a degree-1 solution relatively cheaply.
The resulting velocity field is then used as a much more informed initial guess for a degree-5 solution.
\textcolor{red}{Benchmark.}
Since the vertical basis functions are orthogonal, the projection of the degree-1 solution into the degree-5 function space can be calculated by simply copying array elements.
By contrast, if we tried to use the same strategy to bootstrap a solution with a small number of vertical layers to a solution with a larger number of vertical layers, we would also have to interpolate values to the fine layers.

Finally, we note that this approach was first described in \citet{bassis2010hamilton}.
This work focused solely on flowband models and did not consider the problem of how to calculate the power dissipated at the calving front.
Moreover, this work used only the two vertical basis functions $\psi_1(\zeta) = 1$, $\psi_2(\zeta) = 1 - (1 - \zeta)^{n + 1}$, the idea being that the vertical profile can be written as a direct superposition of the plug flow and shallow ice approximation modes.
By contrast, we leave the number of vertical basis functions up to the user.
This flexibility enables the multigrid-type approach described above.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem via finite differences in the vertical and finite elements in the horizontal, but did not use terrain-following coordinates.
This work also adopts the viewpoint that semi-discretization defines a family of models, rather than merely a problem-specific discretization strategy.

\subsection{Substituting model components}

@@

One of our goals with icepack is to facilitate experimentation with the model physics.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
In C++ one could avoid modifying the diagnostic solve routine by using variadic templates.
Variadic templates are a more advanced language feature and relying on them would present a steep learning curve for novices.

Any model physics component in icepack can be substituted for a parameterization of the user's choosing.
Each flow model object is initialized with a set of functions to calculate the various terms of the action functional.
For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
The default parameterizations are set to the most common choices in the literature, but if the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
Any fields that are unnecessary for calculating a given component are ignored.
For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.

We are restricted to using keyword arguments instead of positional arguments in adopting this approach.
For the particular case of passing several fields to a physics solver routine, using only keyword arguments enhances readability and comprehensibility.
The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
The order of the arguments is arbitrary and immaterial.
For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance and our design would be inappropriate.

@@

\subsection{Convergence criteria} \label{sec:convergence-criteria}

>>>>

\begin{abstract}
In this paper we introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
Icepack is built on the finite element modeling library firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
The diagnostic models in icepack are formulated through action principles that are specified in UFL.
The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
Many post-processing and analysis tasks on simulation results also amount to the evaluation of some functional.
By using a variational formulation of the model physics, the specification of a problem and the analysis of the solution employ the same conceptual vocabulary.
A third advantage of variational principles is that the action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
Icepack features a 3D diagnostic model based on terrain-following coordinates and vertical spectral discretization.
This model resolves both plug- and shear-flow components of horizontal ice flow with a minimum of extra computational expense over 2D, depth-averaged models.
Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
\end{abstract}

@@

Numerical modeling has become an essential part of the workflow of glaciologists across all disciplines.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like.
\end{enumerate}
Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several glacier flow models already exist and are effective in the hands of experts.
These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
Our goal is to make a tool that will be both accessible to non-experts and useful for experts.

The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}.
The Firedrake project began as an outgrowth of FEniCS and both packages implement the same domain-specific language for describing weak forms of PDE.
Icepack improves upon the groundwork laid in VarGlaS in three main respects.
First, icepack includes a simple 3D flow model that uses several features only available in firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
This model will be described in section \S\ref{sec:physics-hybrid-model}.
Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.

@@

\section{Physics}

\subsection{Mass transport}

The PDE describing how the ice thickness changes in time is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
The thickness is represented using continuous Galerkin basis functions.
The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.

The surface elevation is calculated at each timestep of a simulation as
\begin{equation}
    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
\end{equation}
the first case corresponding to grounded ice and the second case corresponding to floating ice.
In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
Most models assume hydrostasy and icepack does as well.
Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.

\subsection{Momentum transport}

There are \textcolor{green}{three} diagnostic models implemented in icepack.
The \emph{shallow ice approximation} (SIA) and shallow stream approximation (SSA) are 2D models describing the depth-averaged velocity of a glacier.
\textcolor{green}{The SIA model describes the deep interior of ice sheets well, where flow velocities are on the order of 0.1 - 10 m/year.
This model assumes that nearly all flow is by vertical shear, which happens when the ice is frozen to the bed or sliding at a small fraction of the surface speed.}
The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
This model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
The \emph{first-order} or \emph{Blatter-Pattyn} (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

\subsubsection{Variational principles}

All of the diagnostic models in icepack are described through \emph{variational} or \emph{action} principles \citep{dukowicz2010consistent}.
Rather than describe the velocity as the weak solution of a nonlinear PDE, an action principle instead states that the velocity minimizes a functional called the action.
The action consists of four terms:
\begin{align}
    & \text{action} = \iint\text{stress} \times \text{strain rate}\ud z\ud x - \int\text{basal friction} \times \text{sliding velocity}\ud x \nonumber \\
    & \quad - \iint\text{surface slope}\times\text{velocity}\ud z\ud x - \iint\text{ocean pressure}\times\text{velocity}\ud z\ud \gamma
    \label{action-functional}
\end{align}
where $\ud z\ud x$ denotes integration over the entire glacier, $\ud x$ denotes integration over the glacier footprint, and $\ud z\ud\gamma$ over the side wall boundary.
The action has units of power (energy/time) and can be related to the rate of decrease of the thermodynamic free energy.
Moreover, the energy lost to viscous and frictional heating can be calculated from the action and its Legendre transform.
The action principle can be viewed as a consequence of the Onsager reciprocity relations for systems near to equilibrium \citep{de2013non}.

@@

\subsubsection{\textcolor{green}{Shallow ice approximation}}

\subsubsection{Shallow stream approximation}

\subsection{Hybrid model} \label{sec:physics-hybrid-model}

The shallow ice and shallow stream models in icepack are completely standard.
The hybrid flow model that we describe here is based on variational principles, terrain-following coordinates, and vertical spectral methods.
Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.

\subsubsection{Terrain-following coordinates}

Rather than use the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
The terrain-following vertical coordinate $\zeta$ is
\begin{equation}
    \zeta = \frac{z - b}{h}
\end{equation}
where $b$ is the ice base.
We can then think of the computational domain as the Cartesian product of a 2D footprint domain $\Omega$ and the unit interval $[0, 1]$.

Both the bed elevation and thickness depend on $x$ and $y$.
As a result, the formula for the horizontal gradient of a field in terrain-following coordinates includes an additional geometric correction factor.
Letting $\nabla_z$ and $\nabla_\zeta$ denote the horizontal gradient with respectively $z$ and $\zeta$ held constant, the chain rule gives us that
\begin{equation}
    \nabla_zq = \nabla_\zeta q + \frac{\partial q}{\partial\zeta}\nabla\zeta,
\end{equation}
where we can calculate the spatial gradient of $\zeta$ as
\begin{equation}
    \nabla\zeta = -h^{-1}\left\{(1 - \zeta)\nabla b + \zeta\nabla s\right\}.
\end{equation}
Likewise, the strain rate of a vector field can be expressed as
\begin{equation}
    \dot\varepsilon_z(u) = \dot\varepsilon_\zeta(u) + \frac{1}{2}\left(u \otimes\nabla\zeta + \nabla\zeta\otimes u\right)
    \label{eq:geometric-correction}
\end{equation}
where $\otimes$ is the tensor product of two vectors.

@@

We also argue that answers expressed in terrain-following coordinates are more intuitive in some respects than in Cartesian coordinates.
At the bed of a grounded glacier, the vertical velocity in Cartesian coordinates is
\begin{equation}
    w = -\dot a_b + u\cdot\nabla b.
\end{equation}
Knowing that a model gives a vertical velocity at the base of a glacier of, say, 10 cm/year, the modeler needs to also know the bed slope and sliding velocity.
In other words, it is not immediately clear whether the vertical velocity is a result of basal mass balance or of geometry without additional information.
By contrast, the vertical velocity $\omega$ in terrain-following coordinates evaluated at $\zeta = 0$ is completely determined by basal mass balance and ice thickness.

\subsubsection{Spectral discretization}

Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
In Elmer/Ice, for example, the user can extend the finite element discretization into a number of vertical layers.
The number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.

The horizontal velocity for many realistic flows is very smooth as a function of depth and this suggests a different approach.
For example, under the plug flow approximation, the horizontal velocity is constant with depth.
Under the shallow ice approximation, the horizontal velocity varies with depth as $1 - (1 - \zeta)^{n + 1}$ where $n = 3$ is the Glen flow law exponent.
This extra information about our solution suggests a modal rather than a nodal discretization strategy.

Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element.
Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
\begin{equation}
    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
\end{equation}
For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.

The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
This approach can be thought of merely as a way to discretize a PDE that has special structure.
Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
The system is then discretized in the horizontal and solved numerically.
In any case, the code is the same regardless of how one views the underlying mathematics.

@@

\subsection{Substituting model components} \label{sec:physics-substitution}

@@

One of our goals with icepack is to facilitate experimentation with the model physics.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
In C++ one could avoid modifying the diagnostic solve routine by using variadic templates.
Variadic templates are a more advanced language feature and relying on them would present a steep learning curve for novices.

Any model physics component in icepack can be substituted for a parameterization of the user's choosing.
Each flow model object is initialized with a set of functions to calculate the various terms of the action functional.
For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
The default parameterizations are set to the most common choices in the literature, but if the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
Any fields that are unnecessary for calculating a given component are ignored.
For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function makes no use of these arguments.
The user can completely change the functional form and add new arguments, with the restriction that arguments are passed by keyword and not positionally.

In adopting this approach, we are restricted to using keyword arguments instead of positional arguments.
Using only keyword arguments enhances readability and comprehensibility for this particular case of invoking a physics solver.
The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used in the literature -- $C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.
The order of the arguments is arbitrary and immaterial.
The preference for argument passing by name is specific to this use case and not universal.
For example, in defining a bilinear form that represents a non-symmetric PDE, the order of the arguments has intrinsic mathematical significance.
In that case using positional arguments would better correspond with the underlying problem.

@@

\subsection{Data assimilation}

Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
The class \texttt{InverseProblem} represents the specification of the inverse problem.
To specify an inverse problem, the user must provide:
\begin{itemize}
    \item the model object and the method that solves the diagnostic equation,
    \item the objective and regularization functionals,
    \item the observed field and the name of the argument to the diagnostic solver,
    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
    \item extra data passed to the diagnostic solver such as boundary conditions.
\end{itemize}
At present, this class assumes that the observed state is always the ice velocity, but in principle the same design would suffice for more complicated inverse problems.
More importantly, the inverse problem is flexible enough to account for the fact that we might have substituted in our own parameterization for the rheology or friction coefficient.
For example, nearly all studies in the literature use an alternate parameterization of the friction coefficient in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
This information is passed to the inverse problem by specifying the names of the keyword arguments for the observed state and the field we wish to observe.

The \texttt{InverseSolver} class is responsible for actually solving the inverse problem.
This class will be described further in section \S\ref{sec:numerics-inverse-solvers}.
The inverse solver is completely agnostic to the particular parameterization of the physics in terms of the inferred field as long as it guarantees positivity.
This feature interacts seamlessly with the ability to modify individual model physics components as described in \S\ref{sec:physics-substitution}.

\subsection{Damage transport}

\subsection{Heat transport}

\subsection{\textcolor{green}{Plumes}}

\section{Numerics}

In this section we describe how various parts of the model physics were implemented in icepack, including both software design and the numerical solution procedures.
From a design perspective, each flow model is represented by a Python class that has three responsibilities.
First, the model class has to have a method that takes in the various fields (ice velocity, thickness, surface elevation, etc.) and returns a symbolic representation of the action functional for that particular model.
Second, the model class must have a method that takes in the input fields and returns an approximate minimizer of the action functional.
The diagnostic solve method amounts to invoking an external Newton solve procedure on the symbolic action functional that the first method calculates.
The Newton solver itself is completely standard but the convergence criterion is not and will be discussed in section \S\ref{sec:convergence-criteria}.
Finally the model must have a method to update the ice thickness from the current value, the ice velocity, and the mass balance rates.
The prognostic solver will be discussed in section \S\ref{sec:prognostic-model}.

The Unified Form Language for specifying weak forms of PDEs contains all of the primitives necessary to express individual terms of the action functional.
These primitives consist of the basic vector calculus operators like the gradient of a field, tensor calculus operations like taking the dot product of two vectors or tensors, scalar functions like the square root or exponential, and symbolic integration over the mesh or its boundary.
For example, the strain rate for a given velocity field $u$ can be written as \texttt{sym(grad(u))}, where the function \texttt{grad} represents the symbolic gradient of a field and \texttt{sym} represents the symmetrization of a rank-2 tensor.

\subsection{Advective transport} \label{sec:prognostic-model}

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the mass transport equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

\subsection{Convex optimization}

The action functional for each diagnostic model is convex, i.e. the second derivative is positive-definite.
\textcolor{red}{Why it's great...}
